Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2015-09-04 pkg/glmertree/paper/treatment_interactions.Rnw
small touch-ups in modeling section -> more to do
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@773 255f4249-2263-4417-b1ea-8498779e27a7

\begin{figure}[!ht]
\setkeys{Gin}{width=0.5\textwidth}
<<exampleglm, fig=TRUE, height=5, width=5>>=
plot(exampleglm)
@
\caption{Example of a linear fixed-effects model for treatment outcomes, based on the artificial motivating dataset ($N=150$). The dot for Treatment~1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
\label{fig:fixedeffects}
\end{figure}

@@

\verb|R| \cite{R14} was used for generation and analysis of all datasets. Two additional R packages were used: \verb|partykit| (version 1.0-2; \citeNP{HothyZeil14,ZeilyHoth08}) for the estimation of GLM trees, and \verb|lme4| (version 1.1-7; \citeNP{BateyMeac12}) for the estimation of random-effects coefficients.

@@

\begin{table}[!htbp]
	\caption{Tree size distributions for GLM and GLMM tree for datasets without treatment-subgroup interactions.}
	\small
	\begin{tabular}{lrrrrrrr}
		\thickline
		&\multicolumn{7}{c}{tree size}\\
		\cline{2-7}
		&   1    &   3   &   5   &   7   &   9   &  11  &   total \\
		\hline
		GLMM tree	& 20625 & 932 & 43 &0&0&0& 21,600\\
			  		& (.96) & (.04) & ($<.01$) &(.00)&(.00)&(.00)&(1.00) \\
   		GLM tree		& 14501  & 4202 & 2013 & 802 & 79 & 3 & 21,600 \\ 
    			& (.67)  & (.19) & (.09) & (.04) & ($<.01$) & ($<.01$) & (1.00)\\ 
		\hline
  		\multicolumn{8}{l}{\textit{Note. }Bracketed values are proportions. Tree sizes are expressed as the total}\\
  		\multicolumn{8}{l}{number of nodes in a tree. A tree with a total of $J$ nodes has $(J+1)/2$ }\\
  		\multicolumn{8}{l}{terminal nodes; the true tree size in datasets without treatment-subgroup}\\	  	\multicolumn{8}{l}{interactions was 1.}\\
	\end{tabular}
	\label{tab:treesize_main}
\end{table}

\begin{figure}[!htbp]
<<treesize-nointeract-xyplot, fig=TRUE, width=9, height=7>>=
load("treesizes_long_nointeract.dat")
treesizes.long$sigmabm <- factor(as.numeric(as.character(treesizes.long$sigmabm)), ordered=T)
treesizes.long$N <- factor(as.numeric(as.character(treesizes.long$N)), ordered=T)
aggdata <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmabm + corUbm, FUN=mean, data=treesizes.long)
print(xyplot(treesize.values ~ sigmabm | N + corUbm, data = aggdata, groups=treesize.ind, 
	type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=FALSE), 
	abline=c(1,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1, 
	lines=TRUE, points=TRUE)))
@
\caption{Average tree size of GLM and GLMM trees for datasets without treatment-subgroup interactions. Values $200$, $500$ and $1000$ refer to sample size. Reference line at $y=1$ represents the true tree size.}
\label{fig:xyplot_treesize_nointeract}
\end{figure}

@@

\begin{table}[!htbp]
\caption{Tree size distributions for GLM and GLMM tree for datasets with treatment-subgroup interactions.}
\small
	\begin{tabular}{lrrrrrrrr}
		\thickline
		&&\multicolumn{7}{c}{tree size}\\
		\cline{2-8}
		&     3   &   5   &   7   &   9   &  11  &   13   &   15  & total \\
		\hline
	  	GLMM tree	& 2 & 281& 29358 & 2675 & 83 & 1 &0& 32,400 \\
					&  ($<.01$)  & ($<.01$)& (.91) & ($.08$) & ($<.01$) & ($<.01$) &(.00)& (1.00) \\

@@

\begin{figure}[!htbp]
<<treesize-xyplot, fig=TRUE, width=9, height=9>>=
aggdata.size <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)	
print(xyplot(treesize.values ~ sigmab | N + corUb, data = aggdata.size, groups=treesize.ind, 
	type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
	abline=c(7,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
)
@
\caption{Average tree size of GLM and GLMM trees for datasets with treatment-subgroup interactions. Values $200$, $500$ and $1000$ refer to sample size. Reference line at $y=7$ represents true tree size.}
\label{fig:xyplot_treesize_interact}
\end{figure}

@@

\begin{figure}[!htbp]
<<accuracy-xyplot, fig=TRUE, width=9, height=9>>=
aggdata.acc <- aggregate(formula=truetree.values ~ truetree.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)
print(xyplot(truetree.values ~ sigmab | N + corUb, data = aggdata.acc, groups=truetree.ind, type="b", 
	ylab="tree accuracy", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=FALSE), 
	auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=TRUE, points=TRUE))
)
@
\caption{Average accuracy of GLM and GLMM trees. Accuracy of trees is defined as the proportion of datasets in which the true tree was accurately recovered. Values are correlated to one of the $U_k$ variables; values $200$, $500$ and $1000$ refer to sample size.}
\label{fig:xyplot_treeaccuracy}
\end{figure}

@@

\begin{figure}[!htbp]
<<correlation-xyplot, fig=TRUE, width=9, height=7>>=
aggdata.cor <- aggregate(formula=correlation.values ~ correlation.ind + N + sigmab + treatdiff, FUN=mean, data=treespecs.long)
print(xyplot(correlation.values ~ sigmab | N + treatdiff, data = aggdata.cor, groups=correlation.ind, 
	type="b", ylab="correlation", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
	auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
)
@
\caption{Average predictive accuracy of GLM and GLMM trees. Predictive accuracy of trees is defined as the correlation between the true and predicted differences between Treatment~1 and~2. Values $5$ and $2.5$ refer to the absolute value of the unstandardized treatment-effect difference in subgroups with treatment-effect differences; values $200$, $500$ and $1000$ refer to sample size.}
\label{fig:xyplot_correlations}
\end{figure}

@@

\begin{figure}[!ht]
\setkeys{Gin}{width=0.75\textwidth}
<<depression-glmtree, fig=TRUE, height=6, width=8>>=
plot(lmtree_app)
@
\caption{GLM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D). The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs. pharmacotherapy (PHA).}
\label{fig:lmtree_C&W}
\end{figure}

@@

\begin{figure}[!ht]
\setkeys{Gin}{width=0.75\textwidth}
<<depression-glmmtree, fig=TRUE, height=4, width=6>>=
plot(lmertree_app$Tree)
@
\caption{GLMM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D). The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs. pharmacotherapy (PHA).}
\label{fig:lmertree_C&W}
\end{figure}

@@

Assessment of predictive accuracy by means of 50-fold cross validation showed that the GLMM tree had higher predictive accuracy than the GLM tree. The correlation between true and predicted posttreatment HAM-D total scores, averaged over the 50 folds, was .28 (var=.067) for GLMM tree, and .19 (var=.084) for GLM tree. This indicates that GLMM tree provided higher predictive accuracy, on average, and also somewhat lower variability of predictive accuracy than GLM tree.

@@

\NumTabs{12}
\noindent $1,\dots,i,\dots,N$ \tab{observation number}\\
$1,\dots,j,\dots,J$ \tab{terminal node number in a tree}\\
$1,\dots,k,\dots,K$ \tab{partitioning variable number}\\
$1,\dots,m,\dots,M$	\tab{cluster number}\\
$\beta_{j}$		\tab\tab{column vector of fixed-effects coefficients in terminal node $j$}\\
$b_{m}$ 		\tab\tab{column vector of random-effects coefficients in cluster $m$}\\
$d_{j}$			\tab\tab{$\beta_{j1} / \sigma_{\epsilon}$; effect size of treatment-effect differences between Treatment~1 and}\\
\tab\tab\tab\tab{Treatment~2 in terminal node $j$}\\
$\epsilon$		\tab\tab{deviation of observed treatment outcome $y$ from its expected value}\\
$r$				\tab\tab{iteration number}\\
$\sigma_{b}$	\tab\tab{square root of variance of $b$}\\
$\sigma_{\epsilon}$\tab\tab{square root of the variance of $\epsilon$}\\
$U_k$			\tab\tab{(potential) partitioning variable $k$}\\
$x_i$ 			\tab\tab{column vector of fixed-effects predictor variable values for observation $i$}\\
$y_i$ 			\tab\tab{treatment outcome for observation $i$}\\
$z_i$ 			\tab\tab{column vector of random-effects predictor variable values for observation $i$}\\

>>>>

\begin{figure}[!th]
\setkeys{Gin}{width=0.5\textwidth}
<<exampleglm, fig=TRUE, height=5, width=5>>=
plot(exampleglm, tp_args = list(mainlab = "Full sample (N = 150)"))
@
\caption{Example of a normal GLM (with fixed effects only) for treatment outcomes, based on the artificial motivating dataset ($N=150$). The dot for Treatment~1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
\label{fig:fixedeffects}
\end{figure}

@@

\verb|R| \cite{R14} was used for generation and analysis of all datasets. Two additional R packages were used: \verb|partykit| (version 1.0-2; \citeNP{HothyZeil15}) for the estimation of GLM trees, and \verb|lme4| (version 1.1-7; \citeNP{BateyMeac12}) for the estimation of random-effects coefficients.

@@

\begin{table}[!thbp]
	\caption{Tree size distributions for GLM and GLMM tree for datasets without treatment-subgroup interactions.}
	\small
	\begin{tabular}{lrrrrrrr}
		\thickline
		&\multicolumn{7}{c}{tree size}\\
		\cline{2-7}
		&   1    &   3   &   5   &   7   &   9   &  11  &   total \\
		\hline
		GLMM tree	& 20625 & 932 & 43 &0&0&0& 21,600\\
			  		& (.96) & (.04) & ($<.01$) &(.00)&(.00)&(.00)&(1.00) \\
   		GLM tree		& 14501  & 4202 & 2013 & 802 & 79 & 3 & 21,600 \\ 
    			& (.67)  & (.19) & (.09) & (.04) & ($<.01$) & ($<.01$) & (1.00)\\ 
		\hline
  		\multicolumn{8}{l}{\textit{Note. }Bracketed values are proportions. Tree sizes are expressed as the total}\\
  		\multicolumn{8}{l}{number of nodes in a tree. A tree with a total of $J$ nodes has $(J+1)/2$ }\\
  		\multicolumn{8}{l}{terminal nodes; the true tree size in datasets without treatment-subgroup}\\	  	\multicolumn{8}{l}{interactions was 1.}\\
	\end{tabular}
	\label{tab:treesize_main}
\end{table}

\begin{figure}[!thbp]
<<treesize-nointeract-xyplot, fig=TRUE, width=9, height=7>>=
load("treesizes_long_nointeract.dat")
treesizes.long$sigmabm <- factor(as.numeric(as.character(treesizes.long$sigmabm)), ordered=T)
treesizes.long$N <- factor(as.numeric(as.character(treesizes.long$N)), ordered=T)
aggdata <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmabm + corUbm, FUN=mean, data=treesizes.long)
print(xyplot(treesize.values ~ sigmabm | N + corUbm, data = aggdata, groups=treesize.ind, 
	type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=FALSE), 
	abline=c(1,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1, 
	lines=TRUE, points=TRUE)))
@
\caption{Average tree size of GLM and GLMM trees for datasets without treatment-subgroup interactions. Values $200$, $500$ and $1000$ refer to sample size. Reference line at $y=1$ represents the true tree size.}
\label{fig:xyplot_treesize_nointeract}
\end{figure}

@@

\begin{table}[!thbp]
\caption{Tree size distributions for GLM and GLMM tree for datasets with treatment-subgroup interactions.}
\small
	\begin{tabular}{lrrrrrrrr}
		\thickline
		&&\multicolumn{7}{c}{tree size}\\
		\cline{2-8}
		&     3   &   5   &   7   &   9   &  11  &   13   &   15  & total \\
		\hline
	  	GLMM tree	& 2 & 281& 29358 & 2675 & 83 & 1 &0& 32,400 \\
					&  ($<.01$)  & ($<.01$)& (.91) & ($.08$) & ($<.01$) & ($<.01$) &(.00)& (1.00) \\

@@

\begin{figure}[!thbp]
<<treesize-xyplot, fig=TRUE, width=9, height=9>>=
aggdata.size <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)	
print(xyplot(treesize.values ~ sigmab | N + corUb, data = aggdata.size, groups=treesize.ind, 
	type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
	abline=c(7,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
)
@
\caption{Average tree size of GLM and GLMM trees for datasets with treatment-subgroup interactions. Values $200$, $500$ and $1000$ refer to sample size. Reference line at $y=7$ represents true tree size.}
\label{fig:xyplot_treesize_interact}
\end{figure}

@@

\begin{figure}[!thbp]
<<accuracy-xyplot, fig=TRUE, width=9, height=9>>=
aggdata.acc <- aggregate(formula=truetree.values ~ truetree.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)
print(xyplot(truetree.values ~ sigmab | N + corUb, data = aggdata.acc, groups=truetree.ind, type="b", 
	ylab="tree accuracy", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=FALSE), 
	auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=TRUE, points=TRUE))
)
@
\caption{Average accuracy of GLM and GLMM trees. Accuracy of trees is defined as the proportion of datasets in which the true tree was accurately recovered. Values are correlated to one of the $U_k$ variables; values $200$, $500$ and $1000$ refer to sample size.}
\label{fig:xyplot_treeaccuracy}
\end{figure}

@@

\begin{figure}[!thbp]
<<correlation-xyplot, fig=TRUE, width=9, height=7>>=
aggdata.cor <- aggregate(formula=correlation.values ~ correlation.ind + N + sigmab + treatdiff, FUN=mean, data=treespecs.long)
print(xyplot(correlation.values ~ sigmab | N + treatdiff, data = aggdata.cor, groups=correlation.ind, 
	type="b", ylab="correlation", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
	auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
)
@
\caption{Average predictive accuracy of GLM and GLMM trees. Predictive accuracy of trees is defined as the correlation between the true and predicted differences between Treatment~1 and~2. Values $5$ and $2.5$ refer to the absolute value of the unstandardized treatment-effect difference in subgroups with treatment-effect differences; values $200$, $500$ and $1000$ refer to sample size.}
\label{fig:xyplot_correlations}
\end{figure}

@@

\begin{figure}[!th]
\setkeys{Gin}{width=0.75\textwidth}
<<depression-glmtree, fig=TRUE, height=6, width=8>>=
plot(lmtree_app)
@
\caption{GLM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D). The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs. pharmacotherapy (PHA).}
\label{fig:lmtree_C&W}
\end{figure}

@@

\begin{figure}[!th]
\setkeys{Gin}{width=0.6\textwidth}
<<depression-glmmtree, fig=TRUE, height=4, width=4.7>>=
plot(lmertree_app$Tree)
@
\caption{GLMM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D). The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs. pharmacotherapy (PHA).}
\label{fig:lmertree_C&W}
\end{figure}

@@

Assessment of predictive accuracy by means of 50-fold cross validation showed that the GLMM tree had higher predictive accuracy than the GLM tree. The correlation between true and predicted posttreatment HAM-D total scores, averaged over the 50 folds, was .28 ($\mathrm{var}=.067$) for GLMM tree, and .19 ($\mathrm{var}=.084$) for GLM tree. This indicates that GLMM tree provided higher predictive accuracy, on average, and also somewhat lower variability of predictive accuracy than GLM tree.

@@

\NumTabs{12}
\noindent $1,\dots,i,\dots,N$ \tab{observation number}\\
$1,\dots,j,\dots,J$	\tab{terminal node number in a tree}\\
$1,\dots,k,\dots,K$	\tab{partitioning variable number}\\
$1,\dots,m,\dots,M$	\tab{cluster number}\\
$\beta_{j}$		\tab\tab{column vector of fixed-effects coefficients in terminal node $j$}\\
$b_{m}$ 		\tab\tab{column vector of random-effects coefficients in cluster $m$}\\
$d_{j}$			\tab\tab{$\beta_{j1} / \sigma_{\epsilon}$; effect size of treatment-effect differences between Treatment~1 and}\\
			\tab\tab\tab\tab{Treatment~2 in terminal node $j$}\\
$\epsilon$		\tab\tab{deviation of observed treatment outcome $y$ from its expected value}\\
$r$			\tab\tab{iteration number}\\
$\sigma_{b}$		\tab\tab{standard deviation of $b$}\\
$\sigma_{\epsilon}$	\tab\tab{standard deviation of $\epsilon$}\\
$U_k$			\tab\tab{(potential) partitioning variable $k$}\\
$x_i$ 			\tab\tab{column vector of fixed-effects predictor variable values for observation $i$}\\
$y_i$ 			\tab\tab{treatment outcome for observation $i$}\\
$z_i$ 			\tab\tab{column vector of random-effects predictor variable values for observation $i$}\\
